---
title: "Data Mining Lab 3"
author: "Gregroy Asamoah, David Churchman, Ann Nelson"
date: "July 28, 2018"
output:
  html_document:
    theme: journal
    highlight: pygments
    number_sections: true
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Business Understanding
<!--[10 points] Describe the purpose of the data set you selected (i.e., why was this data collected in the first place?). How will you measure the effectiveness of a good algorithm? Why does your chosen validation method make sense for this specific dataset and the stakeholders needs?-->

With frequent news of school shootings and police-related violence, there is widespread interest in studying data associated with gun violence incidents. A data set recording all U.S. gun violence events between January 2013 and March 2018, is available for study on [kaggle.com](https://www.kaggle.com/jameslko/gun-violence-data). This data was mined from the Gun Violence Archive using a Python script by James Ko and posted publicly on Kaggle for open investigation. The Gun Violence Archive collects its data by utilizing, "automated queries, manual research through over 2,000 media sources, aggregates, police blotters, police media outlets and other sources daily. Each incident is verified by both initial researchers and secondary validation processes." 

This dataset contains a record of more than 230,000 U.S. gun-related incidents, providing detailed information about the event location both geographic and descriptive, number of participants, number killed or injured, etc. and data related to the shooters and victims (such as age, relationships, count, gender). The dataset also includes a link to the news source from which the data record was obtained. The dataset was created to help close an information-availability gap and make data related to gun violence readily available for analysis, providing opportunity for identifying correlations, developing predictive trend models, and influencing decision-making. 

In the previous two labs, we did exploratory analysis of the data and created algorithms aimed at predicting the level of fatality of a given shooting incident based on the dataset. We used a variety of classification and regression techniques, but in the end, our predictions were only a small amount greater than chance. Though this data set is extensive, it is still limited in quality, and may not provide enough information or reliable enough information to be highly predictive. 

In this lab, rather than attempting predictions, we will look for associations in the dataset, and attempt to create and interpret rules of association which can explain broad trends in gun violence incidents from the last five years. 

# Data understanding 1
<!--[10 points] Describe the meaning and type of data (scale, values, etc.) for each attribute in the data file. Verify data quality: Are there missing values? Duplicate data? Outliers? Are those mistakes? How do you deal with these problems?-->

Blah blah blah

```{r blocknameshowcode}


```

# Data understanding 2
<!-- [10 points]  Visualize the any important attributes appropriately. Important: Provide an interpretation for any charts or graphs.-->

Blah blah blah

```{r graphwithoutshowingcode, echo=FALSE}
plot(pressure)
```

# Modeling and Evaluation 1
<!-- [10 points] 	Train and adjust parameters -->

# Modeling and Evaluation 2
<!-- [10 points] Evaluate and Compare -->

# Modeling and Evaluation 3
<!-- [10 points] Visualize Results -->

# Modeling and Evaluation 4
<!-- [20 points] Summarize the ramifications -->

# Deployment 
<!-- [10 points] Be critical of your performance and tell the reader how you current model might be usable by other parties. Did you achieve your goals? If not, can you reign in the utility of your modeling? How useful is your model for interested parties (i.e., the companies or organizations that might want to use it)? How would your deploy your model for interested parties? What other data should be collected? How often would the model need to be updated, etc.? -->

# Exceptional Work
<!-- [10 points] You have free reign to provide additional analyses or combine analyses. -->








